---
title: "Generation of Figures 1 and 2"
format: html
execute:
  echo: true
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

# Overview

This document generates the two maps used in the manuscript:

- **Figure 1 (Fig1):** Brazilian states classified by geographic region.
- **Figure 2 (Fig2):** Municipal Köppen climate classification and predominant biome.

All final figures are saved in the `figures/` directory in both PNG and TIFF formats.

# Setup

```{r}
# Packages
library(geobr)
library(sf)
library(dplyr)
library(readr)
library(ggplot2)
library(ggspatial)
library(patchwork)

# Output directory
dir.create("figures", recursive = TRUE, showWarnings = FALSE)

# Clean state boundaries from geobr
# These boundaries are reused in Figures 1 and 2.
states_br <- read_state(year = 2020)

# State label positions
state_labels <- states_br |>
  st_point_on_surface()
```

# Figure 1 — Brazilian geographic regions

Figure 1 shows the five official Brazilian geographic regions. State abbreviations are displayed inside each state.

```{r}
# Add geographic region
states_fig1 <- states_br |>
  mutate(
    region = case_when(
      abbrev_state %in% c("AC", "AP", "AM", "PA", "RO", "RR", "TO") ~ "North",
      abbrev_state %in% c("AL", "BA", "CE", "MA", "PB", "PE", "PI", "RN", "SE") ~ "Northeast",
      abbrev_state %in% c("DF", "GO", "MT", "MS") ~ "Central-West",
      abbrev_state %in% c("ES", "MG", "RJ", "SP") ~ "Southeast",
      abbrev_state %in% c("PR", "RS", "SC") ~ "South"
    )
  )

# Create Figure 1
Fig1 <- ggplot(states_fig1) +
  geom_sf(aes(fill = region), color = "white", linewidth = 0.4) +
  geom_sf_text(
    data = state_labels,
    aes(label = abbrev_state),
    size = 4.5,
    fontface = "bold",
    color = "black"
  ) +
  scale_fill_brewer(
    palette = "Set2",
    name = "Region"
  ) +
  annotation_north_arrow(
    location = "bl",
    which_north = "true",
    style = north_arrow_fancy_orienteering
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 13, face = "bold"),
    legend.text = element_text(size = 12)
  )

Fig1
```

## Save Figure 1

```{r}
ggsave(
  "figures/Fig1.png",
  Fig1,
  width = 10,
  height = 8,
  units = "in",
  dpi = 600,
  bg = "transparent"
)

ggsave(
  "figures/Fig1.tiff",
  Fig1,
  width = 10,
  height = 8,
  units = "in",
  dpi = 600,
  device = "tiff",
  compression = "lzw",
  bg = "transparent"
)
```

# Figure 2 — Köppen climate classification and predominant biome

Figure 2 combines two municipal-level thematic maps:

- **Panel A:** Köppen climate classification.
- **Panel B:** Predominant biome.

Municipal polygons are dissolved by thematic class before plotting. This removes internal municipal border artifacts. Clean state boundaries from `geobr` are then overlaid in black.

```{r}
# Read determinant data and municipal geometry
determinants <- read_csv2("data/banco_determinantes.csv")

municipalities <- st_read(
  "data/muni_br.gpkg",
  quiet = TRUE
) |>
  st_as_sf(sf_column_name = "geom")

# Standardize join variables
determinants <- determinants |>
  mutate(
    muni_code = as.numeric(muni_code),
    koppen = as.character(koppen),
    bioma = as.character(bioma)
  )

municipalities <- municipalities |>
  mutate(code_muni = as.numeric(code_muni))

# Join determinants to municipal polygons
municipal_map <- municipalities |>
  left_join(
    determinants,
    by = c("code_muni" = "muni_code")
  ) |>
  st_make_valid() |>
  mutate(
    koppen_label = factor(
      koppen,
      levels = c("Af", "Am", "Aw", "As", "BSh", "Cfa", "Cfb", "Cwa", "Cwb"),
      labels = c(
        "Af - Tropical rainforest",
        "Am - Tropical monsoon",
        "Aw - Tropical savanna",
        "As - Tropical dry summer",
        "BSh - Hot semi-arid",
        "Cfa - Humid subtropical",
        "Cfb - Oceanic",
        "Cwa - Dry-winter humid subtropical",
        "Cwb - Subtropical highland"
      )
    ),
    biome_label = case_when(
      bioma %in% c("Amazon", "Amazônia", "Amazonia") ~ "Amazon",
      bioma == "Caatinga" ~ "Caatinga",
      bioma == "Cerrado" ~ "Cerrado",
      bioma %in% c("Atlantic Forest", "Mata Atlântica", "Mata Atlantica") ~ "Atlantic Forest",
      bioma == "Pampa" ~ "Pampa",
      bioma == "Pantanal" ~ "Pantanal",
      TRUE ~ bioma
    ),
    biome_label = factor(
      biome_label,
      levels = c(
        "Amazon",
        "Caatinga",
        "Cerrado",
        "Atlantic Forest",
        "Pampa",
        "Pantanal"
      )
    )
  )

# Dissolve municipalities by thematic class
koppen_data <- municipal_map |>
  filter(!is.na(koppen_label)) |>
  group_by(koppen_label) |>
  summarise(.groups = "drop")

biome_data <- municipal_map |>
  filter(!is.na(biome_label)) |>
  group_by(biome_label) |>
  summarise(.groups = "drop")
```

## Common map theme

```{r}
map_theme <- theme_void() +
  theme(
    plot.background = element_rect(
      fill = "transparent",
      color = NA
    ),
    panel.background = element_rect(
      fill = "transparent",
      color = NA
    ),
    legend.background = element_rect(
      fill = "transparent",
      color = NA
    ),
    legend.box.background = element_rect(
      fill = "transparent",
      color = NA
    ),
    plot.title = element_text(
      size = 15,
      face = "bold",
      hjust = 0.5
    ),
    legend.title = element_text(
      size = 10,
      face = "bold"
    ),
    legend.text = element_text(
      size = 8
    ),
    legend.position = "bottom"
  )
```

## Panel A — Köppen climate type

```{r}
koppen_plot <- ggplot() +
  geom_sf(
    data = koppen_data,
    aes(fill = koppen_label),
    color = NA
  ) +
  geom_sf(
    data = states_br,
    fill = NA,
    color = "black",
    linewidth = 0.35
  ) +
  geom_sf_text(
    data = state_labels,
    aes(label = abbrev_state),
    size = 2.8,
    fontface = "bold"
  ) +
  scale_fill_brewer(
    palette = "Paired",
    name = "Köppen climate type"
  ) +
  guides(
    fill = guide_legend(ncol = 2, byrow = TRUE)
  ) +
  annotation_north_arrow(
    location = "tr",
    which_north = "true",
    style = north_arrow_fancy_orienteering
  ) +
  labs(title = "A - Köppen climate type") +
  map_theme
```

## Panel B — Predominant biome

```{r}
biome_plot <- ggplot() +
  geom_sf(
    data = biome_data,
    aes(fill = biome_label),
    color = NA
  ) +
  geom_sf(
    data = states_br,
    fill = NA,
    color = "black",
    linewidth = 0.35
  ) +
  geom_sf_text(
    data = state_labels,
    aes(label = abbrev_state),
    size = 2.8,
    fontface = "bold"
  ) +
  scale_fill_brewer(
    palette = "Set3",
    name = "Predominant biome"
  ) +
  guides(
    fill = guide_legend(ncol = 2, byrow = TRUE)
  ) +
  annotation_north_arrow(
    location = "tr",
    which_north = "true",
    style = north_arrow_fancy_orienteering
  ) +
  labs(title = "B - Predominant biome") +
  map_theme
```

## Combine and save Figure 2

```{r}
Fig2 <- koppen_plot + biome_plot +
  plot_layout(ncol = 2)

Fig2

ggsave(
  "figures/Fig2.png",
  Fig2,
  width = 16,
  height = 8,
  units = "in",
  dpi = 600,
  bg = "transparent"
)

ggsave(
  "figures/Fig2.tiff",
  Fig2,
  width = 16,
  height = 8,
  units = "in",
  dpi = 600,
  device = "tiff",
  compression = "lzw",
  bg = "transparent"
)
```
