---
title: "Figure 4. Dengue transmission patterns in Brazil"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    theme: cosmo
editor: visual
execute:
  warning: false
  message: false
editor_options:
  chunk_output_type: console
---

## Objective

This script creates a two-panel map of dengue transmission patterns in Brazil for two periods:

- **A - 2010–2016**
- **B - 2017–2022**

This version was simplified to avoid geometry artifacts. It uses the **municipal mesh only for the colored areas** and overlays **clean state boundaries** from `geobr`. Municipal border lines are removed. The final figure uses **one shared horizontal legend**.

## Packages

```{r}
# Install if needed:
# install.packages(c("sf", "dplyr", "readr", "ggplot2", "patchwork", "stringr", "grid", "geobr"))

library(sf)
library(dplyr)
library(readr)
library(ggplot2)
library(patchwork)
library(stringr)
library(grid)
library(geobr)
```

## Helper function to find files

```{r}
find_file <- function(candidates) {
  found <- candidates[file.exists(candidates)]
  
  if (length(found) == 0) {
    stop(
      "File not found. I tried these paths:\n",
      paste(candidates, collapse = "\n")
    )
  }
  
  found[1]
}

file_10_16 <- find_file(c(
  "table/2010-2016/dengue_pattern_10_16.csv",
  "tables/2010-2016/dengue_pattern_10_16.csv",
  "Tabelas_dengue/2010-2016/dengue_pattern_10_16.csv",
  "../table/2010-2016/dengue_pattern_10_16.csv",
  "../tables/2010-2016/dengue_pattern_10_16.csv",
  "../Tabelas_dengue/2010-2016/dengue_pattern_10_16.csv"
))

file_17_22 <- find_file(c(
  "table/2017-2022/dengue_pattern_17_22.csv",
  "tables/2017-2022/dengue_pattern_17_22.csv",
  "Tabelas_dengue/2017-2022/dengue_pattern_17_22.csv",
  "../table/2017-2022/dengue_pattern_17_22.csv",
  "../tables/2017-2022/dengue_pattern_17_22.csv",
  "../Tabelas_dengue/2017-2022/dengue_pattern_17_22.csv"
))

file_10_16
file_17_22
```

## Read municipal and state meshes

`geobr` is used here to obtain clean geometries and avoid the visual artifacts that appeared before.

```{r}
municipalities <- read_municipality(code_muni = "all", year = 2020, simplified = TRUE)
states_br <- read_state(code_state = "all", year = 2020, simplified = TRUE)

municipalities <- municipalities %>%
  mutate(code_muni = as.numeric(code_muni))
```

## Read dengue pattern tables

```{r}
padrao_10_16 <- read_csv2(file_10_16) %>%
  mutate(
    muni_code = as.numeric(muni_code),
    dengue_pattern = as.character(dengue_pattern)
  )

padrao_17_22 <- read_csv2(file_17_22) %>%
  mutate(
    muni_code = as.numeric(muni_code),
    dengue_pattern = as.character(dengue_pattern)
  )
```

## Join pattern tables to municipalities

```{r}
mapa_10_16 <- municipalities %>%
  left_join(padrao_10_16, by = c("code_muni" = "muni_code"))

mapa_17_22 <- municipalities %>%
  left_join(padrao_17_22, by = c("code_muni" = "muni_code"))
```

## Translate and order classes

```{r}
translate_pattern <- function(x) {
  x <- str_trim(x)
  
  case_when(
    x %in% c(
      "Transmissão Persistente",
      "Transmissao Persistente",
      "transmissão persistente",
      "transmissao persistente"
    ) ~ "Persistent transmission",
    
    x %in% c(
      "Epidêmico",
      "Epidemico",
      "epidêmico",
      "epidemico"
    ) ~ "Epidemic",
    
    x %in% c(
      "Episódico/Epidêmico",
      "Episodico/Epidemico",
      "episódico/epidêmico",
      "episodico/epidemico"
    ) ~ "Episodic/Epidemic",
    
    x %in% c(
      "Episódico",
      "Episodico",
      "episódico",
      "episodico"
    ) ~ "Episodic",
    
    x %in% c(
      "Sem transmissão",
      "Sem transmissao",
      "sem transmissão",
      "sem transmissao",
      "Sem detecção de transmissão",
      "Sem deteccao de transmissao",
      "sem detecção de transmissão",
      "sem deteccao de transmissao"
    ) ~ "No transmission",
    
    TRUE ~ NA_character_
  )
}

legend_order <- c(
  "Persistent transmission",
  "Epidemic",
  "Episodic/Epidemic",
  "Episodic",
  "No transmission"
)

mapa_10_16 <- mapa_10_16 %>%
  mutate(pattern_en = factor(translate_pattern(dengue_pattern), levels = legend_order))

mapa_17_22 <- mapa_17_22 %>%
  mutate(pattern_en = factor(translate_pattern(dengue_pattern), levels = legend_order))
```

## Colors

```{r}
pattern_colors <- c(
  "Persistent transmission" = "#F4C430",
  "Epidemic" = "#F6E58D",
  "Episodic/Epidemic" = "#2E86C1",
  "Episodic" = "#76D7C4",
  "No transmission" = "#6E6E6E"
)
```

## Common theme

```{r}
map_theme <- theme_void() +
  theme(
    plot.background = element_rect(fill = "transparent", color = NA),
    panel.background = element_rect(fill = "transparent", color = NA),
    legend.background = element_rect(fill = "transparent", color = NA),
    legend.box.background = element_rect(fill = "transparent", color = NA),
    plot.title = element_text(size = 15, face = "bold", hjust = 0.5),
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8),
    legend.position = "bottom"
  )
```

## Manual north arrow

```{r}
make_north_arrow <- function(bbox) {
  x <- bbox["xmax"] - 4.0
  y <- bbox["ymax"] - 3.5
  
  list(
    annotate(
      "segment",
      x = x, xend = x,
      y = y - 1.0, yend = y + 0.4,
      arrow = arrow(length = unit(0.22, "cm"), type = "closed"),
      linewidth = 0.5,
      color = "black"
    ),
    annotate(
      "text",
      x = x,
      y = y + 0.8,
      label = "N",
      fontface = "bold",
      size = 4
    )
  )
}

bbox_br <- st_bbox(states_br)
north_arrow <- make_north_arrow(bbox_br)
```

## Shared fill scale

Using the same scale object in both panels helps patchwork collect **one single legend**.

```{r}
fill_scale <- scale_fill_manual(
  values = pattern_colors,
  breaks = legend_order,
  drop = FALSE,
  na.value = "white",
  name = "Transmission pattern"
)
```

## Panel A

```{r}
map_10_16_plot <- ggplot() +
  geom_sf(
    data = mapa_10_16,
    aes(fill = pattern_en),
    color = NA,
    linewidth = 0
  ) +
  geom_sf(
    data = states_br,
    fill = NA,
    color = "black",
    linewidth = 0.30
  ) +
  north_arrow +
  fill_scale +
  labs(title = "A - 2010–2016") +
  guides(fill = guide_legend(nrow = 1, byrow = TRUE)) +
  coord_sf(expand = FALSE) +
  map_theme
```

## Panel B

```{r}
map_17_22_plot <- ggplot() +
  geom_sf(
    data = mapa_17_22,
    aes(fill = pattern_en),
    color = NA,
    linewidth = 0
  ) +
  geom_sf(
    data = states_br,
    fill = NA,
    color = "black",
    linewidth = 0.30
  ) +
  north_arrow +
  fill_scale +
  labs(title = "B - 2017–2022") +
  guides(fill = guide_legend(nrow = 1, byrow = TRUE)) +
  coord_sf(expand = FALSE) +
  map_theme
```

## Combine maps with one shared legend

```{r, fig-width=16, fig-height=8}
Fig4 <- map_10_16_plot + map_17_22_plot +
  plot_layout(ncol = 2, guides = "collect") &
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8)
  )

Fig4
```

## Save figure

```{r}
dir.create("figures", recursive = TRUE, showWarnings = FALSE)

ggsave(
  "figures/Fig4.png",
  plot = Fig4,
  width = 16,
  height = 8,
  units = "in",
  dpi = 300,
  bg = "white"
)

ggsave(
  "figures/Fig4.tiff",
  plot = Fig4,
  width = 16,
  height = 8,
  units = "in",
  dpi = 600,
  device = "tiff",
  compression = "lzw",
  bg = "white"
)
```

## Suggested caption

**Fig. 4. Classification of Brazilian municipalities according to dengue transmission patterns.** A: 2010–2016. B: 2017–2022.
